---
title: "StatInference_Project - Part B"
author: "Georgia P"
date: "September 14, 2015"
output: html_document
---
#Overview
In this analysis the goal is analyze growth data for teeth (ToothGrowth Data) and explore statistical significance of evolution measured as **Tooth Length** (len variable) as well as the correlation with parameters such as **Supplement type** (VC or OJ) (supp variable) and **Dose in milligrams** (dose variable).

#Exploratory analysis
```{r echo=TRUE, fig.height= 9, message=FALSE} 
library(datasets) ; library(ggplot2) ; library(gridExtra)
tdata<-ToothGrowth
#basic summary of dataset
summary(tdata)

# scatterplot/  primary visualization of my dataset
sq<- qplot(data=tdata, y=len,x=dose, color= supp, xlab = "dose category", ylab="teeth length",main="Teeth lengths per dose and supplement types", labels="supplement type")

# boxplot/ informative visualization of my variables statistics
bq<- qplot(data=tdata, y=len, x= supp,fill=supp, facets=. ~ dose,geom="boxplot", xlab = "dose category",ylab="teeth length", main="Length boxplots per dose and supplement types", labels="supplement type")

# plot variable distributions/ densities, better understand my variable characteristics
dq<- qplot(data=tdata, x=len, color= supp,geom="density",  facets=. ~ dose, xlab = "length | dose category", main="Length density per dose and supplement types", labels="supplement type")

#plot arranged
grid.arrange(sq, bq,dq, nrow=3)

```

# Statistical Inference 

## Hypothesis
Exploring the available data, I observe that lengths of teeth overlap among supplement types and dose categories. Therefore, because teeth length is an indicator that can only grow, not shrink back to original size for a second iteration (:)), I assume that there were at lease 6 **independent subject groups**, each one of them treated with one dose category (0.5mm /1mm/ 2mm) and one type of supplement (OJ / VJ).
```{r fig.height= 4.5,echo=TRUE }
#make classes and arrange
library(dplyr)
tdata2<- mutate(tdata, class=paste(supp, dose, "mm")) %>% group_by(class) %>% arrange(len)
qplot(data=tdata2, y=len,color= class, xlab = "index", ylab="teeth length",main="Teeth lengths per class (dose & supplement types)")
```


Some criteria that you will be evaluated on
Did you  perform an exploratory data analysis of at least a single plot or table highlighting basic features of the data?
Did the student perform some relevant confidence intervals and/or tests?
Were the results of the tests and/or intervals interpreted in the context of the problem correctly? 
Did the student describe the assumptions needed for their conclusions?